---
title: "Food Tree (Figure 1D)"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)   # data wrangling + readr/tidyr/dplyr/stringr
library(data.tree)   # build hierarchical tree from paths
library(phytools)    # read.newick() for sanity-check loading
library(here)        # project-root-relative paths
```

# Overview

This notebook builds a circular food taxonomy tree (Figure 1D) from the
combined diet table, using GraPhlAn for the final rendering.

The pipeline is:

1.  Build a 7-level food taxonomy from each `Food_code` (one digit per
    level).
2.  Write the taxonomy out as a Newick tree (one for IDs, one for
    names).
3.  Build GraPhlAn annotation files (per-meal average dehydrated weight
    becomes the outer ring; food group → ring color).
4.  Call GraPhlAn (via Docker) to render the PDF.

Code adapted from <https://github.com/knights-lab/Food_Tree>.

## Sources for the taxonomy levels

The level labels are part USDA and part not. Both citations, and the split
level by level, are in [Where the level labels come from] below.

1.  U.S. Department of Agriculture, Agricultural Research Service. *USDA Food
    and Nutrient Database for Dietary Studies 2011-2012, Documentation*,
    Appendix E, "Food/Beverage Coding Scheme", pp. 58-72. Food Surveys Research
    Group, 2014.
    <https://www.ars.usda.gov/ARSUserFiles/80400530/pdf/fndds/fndds_2011_2012_doc.pdf>
    Levels 1 to 3, and 67 of the level-4 codes.
2.  Johnson AJ, Vangay P, Al-Ghalith GA, et al. Daily sampling reveals
    personalized diet-microbiome associations in humans. *Cell Host Microbe*
    2019;25(6):789-802.e5. doi:10.1016/j.chom.2019.05.005. Label file at
    <https://github.com/knights-lab/Food_Tree/blob/master/R/data/NodeLabelsMCT.txt>
    The remaining level-4 codes and all of levels 5 and 6.

# Inputs / outputs

```{r paths}
# One data root, resolved from NUTRITION_DATA (falls back to released_data so a
# public clone works with no setup). Regenerable artifacts land in intermediate_data,
# the final panel in results.
data_root <- Sys.getenv("NUTRITION_DATA", unset = here::here("released_data"))
inter_dir <- here::here("intermediate_data")
res_dir   <- here::here("results")

# ---- inputs (released_data) ----
dtb_path        <- file.path(data_root, "152_combined_DTB.csv")           # diet table (per-portion records)
nodes_path      <- file.path(data_root, "NodeLabelsMCT.txt")              # level code -> label, plus provenance columns
fg_color_path   <- file.path(data_root, "food_group_color_key_final.csv") # food-group -> ring color
annot_base_path <- file.path(data_root, "annotation.base.txt")            # GraPhlAn global style header

# ---- outputs (intermediate_data) ----
out_table_csv   <- file.path(inter_dir, "final_table_for_writing_out_to_newick.csv")
out_tree_id     <- file.path(inter_dir, "output_food_tree_datatree.newick")      # leaves = Food_code
out_tree_name   <- file.path(inter_dir, "output_food_tree_datatree_name.newick") # leaves = description (GraPhlAn input)
ring_annot_tsv  <- file.path(inter_dir, "ring_annot.txt")
ring_style_tsv  <- file.path(inter_dir, "ring_itself.txt")

# Food_code is 8 digits. L1..L7 are built from the first 1..7 digits, but only
# L1..L6 become tree levels; the leaf is the food code itself, so L7 is carried
# in the taxonomy string only.
N_LEVELS <- 7
```

# 1. Load the food database

Take all unique `(Food_code, description)` pairs from the diet table and
sanitize the description (Newick / GraPhlAn dislike spaces, commas,
parentheses, semicolons, slashes, quotes).

```{r load-foods}
dtb <- read_csv(dtb_path)

food_database <- dtb %>%
  distinct(Food_code, description) %>%
  distinct(Food_code, .keep_all = TRUE) %>%   # guard against same code w/ multiple descriptions
  rename(FoodID = Food_code) %>%
  mutate(
    description = description %>%
      str_replace_all("[ ,\\-()%/\";]", "_") %>%
      str_remove_all("'")
  )
```

# 2. Build the 7-level taxonomy table

Each digit position of `FoodID` is a taxonomy level. We look up each
level code in the MCT node table to attach a human-readable label.

```{r build-taxonomy}
# The label file also carries provenance columns (see the next section); only the
# code and its label take part in the taxonomy build.
node_labels <- read_tsv(nodes_path, col_types = cols(.default = "c"))
nodes <- node_labels %>% select(Level.code, Main.food.description)

# Build columns L1..L7 by progressively taking the first 1..N_LEVELS digits of FoodID.
main <- food_database %>%
  select(FoodID, description)

for (k in seq_len(N_LEVELS)) {
  main[[paste0("L", k)]] <- stringr::str_sub(main$FoodID, 1, k)
}

# Replace each level code with its human-readable label from `nodes`.
main_named <- main %>%
  pivot_longer(starts_with("L"), names_to = "Level", values_to = "Level.code") %>%
  left_join(nodes, by = "Level.code") %>%
  select(-Level.code) %>%
  pivot_wider(names_from = Level, values_from = Main.food.description) %>%
  mutate(across(starts_with("L"), ~ replace_na(.x, "")))

# Newick path string (leaves = FoodID, internal nodes = level labels).
final_table <- main_named %>%
  mutate(
    newickstring      = str_glue("foodtreeroot/{L1}/{L2}/{L3}/{L4}/{L5}/{L6}/{FoodID}"),
    newickstring_name = str_glue("foodtreeroot/L1_{L1}/L2_{L2}/L3_{L3}/L4_{L4}/L5_{L5}/L6_{L6}/{description}"),
    taxonomy          = pmap_chr(across(starts_with("L")), ~ paste(c(...), collapse = ";"))
  )

dir.create(dirname(out_table_csv), showWarnings = FALSE, recursive = TRUE)
write_csv(final_table, out_table_csv)
```

## Where the level labels come from

`NodeLabelsMCT.txt` is not a USDA file. Its first two columns come unchanged from
`R/data/NodeLabelsMCT.txt` in the Knights lab food-tree repository
(<https://github.com/knights-lab/Food_Tree/blob/master/R/data/NodeLabelsMCT.txt>),
the code base behind Johnson AJ, Vangay P, Al-Ghalith GA, et al. *Daily sampling
reveals personalized diet-microbiome associations in humans.* Cell Host Microbe
2019;25(6):789-802.e5. doi:10.1016/j.chom.2019.05.005. The remaining four columns
are added here and record where each label came from.

Only part of it is the official USDA food coding scheme:

- Levels 1 to 3 (the first 1, 2 and 3 digits of the food code) are USDA
  subgroups, published as Appendix E, "Food/Beverage Coding Scheme", pp. 58-72 of
  the FNDDS 2011-2012 documentation
  (<https://www.ars.usda.gov/ARSUserFiles/80400530/pdf/fndds/fndds_2011_2012_doc.pdf>).
  That is the last FNDDS release to publish the scheme below two digits. Current
  documentation (FNDDS 2021-2023, Appendix H) publishes the first two digits only.
- Level 4 is official for the 67 four-digit subgroups USDA actually defines
  (mostly meat/poultry/fish mixtures and beverages). The other 330 four-digit
  labels are not USDA.
- Levels 5 and 6 are entirely a Food_Tree extension: 765 five-digit and 27
  six-digit labels with no USDA counterpart, read off the FNDDS food descriptions
  by the original authors.

The label file tracks the FNDDS 2011-2012 scheme specifically: the only official
codes missing from it (415, 2812, 9256, 9257, 9265) are exactly the ones that
release discontinued or moved, and the ones it adds at level 3 and below (95x
formulated nutrition beverages, 537x bars) are the ones that release added.

The added columns are `n_digits`, `source` (`usda_fndds` or
`food_tree_extension`), `usda_fndds_description` (USDA's own wording, where the
code is official) and `used_in_F1d_tree` (whether any food code in this cohort
reaches that label, since a label only shapes the tree if some food hits it). The
USDA wording is a one-time transcription of the Appendix E table above, not
re-scraped at run time.

```{r provenance-summary}
# Sanity check: every label carries a source, and the level number matches the code.
stopifnot(
  all(node_labels$source %in% c("usda_fndds", "food_tree_extension")),
  all(as.integer(node_labels$n_digits) == nchar(node_labels$Level.code))
)

node_labels %>%
  count(n_digits, source, used_in_F1d_tree) %>%
  pivot_wider(names_from = c(source, used_in_F1d_tree), values_from = n, values_fill = 0)
```

# 3. Write Newick trees

Two trees are produced from the same taxonomy: one with `Food_code`
leaves (used downstream for joining numeric data) and one with
`description` leaves (used by GraPhlAn so labels are readable).

```{r write-newick}
write_newick <- function(df, path_col, out_path) {
  tree_obj <- data.tree::as.Node(df, pathName = path_col)
  dir.create(dirname(out_path), showWarnings = FALSE, recursive = TRUE)
  cat(data.tree::ToNewick(tree_obj), file = out_path)
}

write_newick(final_table, "newickstring",      out_tree_id)
write_newick(final_table, "newickstring_name", out_tree_name)
```

# 4. Per-food summaries for the outer ring

Outer-ring height = average dehydrated weight per meal for each food.
Outer-ring color = food-group color (first digit of `Food_code`).

```{r ring-data}
# Total number of distinct meals across the whole cohort (denominator).
n_meal <- dtb %>%
  distinct(pid, fdrt, Meal) %>%
  nrow()

# Tree leaves, in tree order — we need our ring rows to match this order.
leaves <- read.newick(out_tree_id)$tip.label

# Food-group color key, e.g. fgrp1 = "1" → fg_milk → "#..."
fg_colors <- read_csv(fg_color_path, col_types = "ccccc")

# Custom ordering for the food-group factor (drives the ring color sweep).
fg_levels <- c("fg_egg", "fg_oils", "fg_legume", "fg_fruit", "fg_sweets",
               "fg_milk", "fg_meat", "fg_veggie", "fg_grain")

ring_data <- dtb %>%
  group_by(Food_code) %>%
  summarise(fc_sum = sum(dehydrated_weight), .groups = "drop") %>%
  mutate(Food_code = as.character(Food_code)) %>%
  filter(Food_code %in% leaves) %>%
  mutate(
    per_meal_ave = round(fc_sum / n_meal, 3),
    fgrp1        = str_sub(Food_code, 1, 1)
  ) %>%
  left_join(fg_colors, by = "fgrp1") %>%
  mutate(fg1_name = factor(fg1_name, levels = fg_levels)) %>%
  arrange(fg1_name)
```

# 5. GraPhlAn annotation files

GraPhlAn expects a long-format TSV with columns
`<leaf_label>  <key>  <ring_level>  <value>`. Leaf labels here are food
*descriptions*, because that's what the `_name.newick` tree carries.

```{r ring-annotation}
# Map Food_code → sanitized description (the leaf label in the _name tree).
fcode_to_desc <- final_table %>%
  select(FoodID, description) %>%
  mutate(FoodID = as.character(FoodID))

# One row per (food, attribute): height, color, width.
ring_annot <- ring_data %>%
  select(Food_code, ring_height = per_meal_ave, ring_color = color) %>%
  left_join(fcode_to_desc, by = c("Food_code" = "FoodID")) %>%
  mutate(
    ring_width  = 0.5,
    # cast all ring attributes to character so pivot_longer can stack them
    across(c(ring_height, ring_color, ring_width), as.character)
  ) %>%
  pivot_longer(c(ring_height, ring_color, ring_width),
               names_to = "key", values_to = "value") %>%
  mutate(ring_level = 1L) %>%
  select(description, key, ring_level, value)

dir.create(dirname(ring_annot_tsv), showWarnings = FALSE, recursive = TRUE)
write_tsv(ring_annot, ring_annot_tsv, col_names = FALSE)
```

```{r ring-style}
# Per-ring styling (separators + label font). Single ring → one row per attribute.
ring_style <- tribble(
  ~key,                                 ~value,
  "ring_internal_separator_thickness",  "0",
  "ring_external_separator_thickness",  "0",
  "ring_label_font_size",               "5",
  "ring_separator_color",               "black",
) %>%
  mutate(ring_level = 1L) %>%
  select(key, ring_level, value)

write_tsv(ring_style, ring_style_tsv, col_names = FALSE)
```

# 6. Concatenate annotations and render with GraPhlAn

`Annot_tree4.txt` = global header (`annotation.base.txt`, the food-group
sector colors and labels) + per-ring style (`ring_itself.txt`) + per-leaf
ring values (`ring_annot.txt`, one food's per-meal-average height/color).

GraPhlAn runs only under Python 2.7, so the simplest reproducible path is
the prebuilt `shengwei/graphlan` Docker image rather than a local install.
The commands below were run and verified to produce the published panel.

**Prerequisites:** Docker Desktop running. On Apple Silicon the
`--platform linux/amd64` flag is required (you may see a harmless platform
warning and a one-time matplotlib font-cache message).

```{bash combine, eval = FALSE}
# Run from the project root. Build the combined annotation file.
cat released_data/annotation.base.txt \
    intermediate_data/ring_itself.txt \
    intermediate_data/ring_annot.txt \
  > intermediate_data/Annot_tree4.txt
```

```{bash render, eval = FALSE}
# Mount intermediate_data as /data inside the container.
# 1) annotate the (description-leaf) tree with the rings -> guide xml
# 2) xml -> final circular PDF
D="$(pwd)/intermediate_data"

docker run --rm --platform linux/amd64 -v "$D":/data shengwei/graphlan:latest \
  graphlan_annotate.py \
    --annot /data/Annot_tree4.txt \
    /data/output_food_tree_datatree_name.newick \
    /data/guide_1_new.xml

docker run --rm --platform linux/amd64 -v "$D":/data shengwei/graphlan:latest \
  graphlan.py \
    /data/guide_1_new.xml \
    /data/F1d_food_tree.pdf \
    --dpi 300 --size 2.5 --pad 0

cp intermediate_data/F1d_food_tree.pdf results/F1d_food_tree.pdf
```

> See <https://huttenhower.sph.harvard.edu/graphlan/> for GraPhlAn notes,
> and <https://hub.docker.com/r/shengwei/graphlan> for the image.

## Figure legend (panel F1d)

> **Figure 1d.** Circular food taxonomy of every food eaten in the cohort.
> Each leaf is one FNDDS food code; internal nodes are the six taxonomy
> levels taken from the first one to six digits of that code. Levels 1 to 3
> are USDA FNDDS subgroups (FNDDS 2011-2012 documentation, Appendix E);
> levels 4 to 6 are mostly the finer split distributed with the Knights lab
> food-tree code (Johnson et al., Cell Host Microbe 2019), which USDA does
> not publish. Sectors are colored by major food group (Grains, Vegetables,
> Meats, Sweets, Milk, Fruits, Legumes, Oils, Eggs) using the shared
> `food_group_color_key_final.csv` palette. The outer ring bar height is
> the average dehydrated weight of that food per meal across all
> `r format(9419, big.mark = ",")` meals (the "Avg. per-meal intake"
> arrow).

The "Avg. per-meal intake" arrow label and the radial scale bar in the
published panel are post-hoc Illustrator annotations on top of the
GraPhlAn ring; everything data-bearing (the tree, sector colors, ring
heights) is produced by this script.

## What to ship for a no-R render

The four inputs above (`152_combined_DTB.csv`, `NodeLabelsMCT.txt`,
`food_group_color_key_final.csv`, `annotation.base.txt`) already in
`released_data/` are enough to regenerate everything. To let a reader
render the tree with Docker alone (skipping the R steps 1-5), also ship the
single essential intermediate, the GraPhlAn input tree:

- `output_food_tree_datatree_name.newick` (description-leaf Newick)

together with the regenerated `ring_annot.txt` / `ring_itself.txt`. The
`final_table_for_writing_out_to_newick.csv` is an optional convenience copy
of the full taxonomy table.



